# Literature {#litreview}

```{r setup, file = "R/chapter_start.R", include = FALSE, cache = FALSE}
# a number of commands need to run at the beginning of each chapter. This
# includes loading libraries that I always use, as well as options for 
# displaying numbers and text.
library(readxl)
```

In a groundbreaking theoretical article, Berdica [@berdica2002] attempted to
identify, define and conceptualize network "vulnerability" --- the complement of
resiliency --- by envisioning analyses conducted with several vulnerability
performance measures including travel time, delay, congestion, serviceability,
and accessibility. She then defined reliability as the level of reduced
accessibility due to unfavorable operating conditions on the network. In
particular, the author identifies a need for further research toward developing
a framework capable of investigating reliability of transportation networks.

In this section we examine several attempts by numerous researchers to do
precisely this using various measures of network performance. It is helpful to
categorize the existing literature into three groups (summarized in
@tbl-authortable) based on the overall technique applied in the study. These
groups include:

-   *Network connectivity*: How does damage to a network diminish the
    connectivity between network nodes?

-   *Travel time analysis*: How much do shortest path travel times between
    origins and destinations increase on a damaged network?

-   *Accessibility analysis* How easily can the population using the damaged
    network complete their daily activities?


```{r authortable, echo = FALSE, cache = FALSE}
#| label: tbl-authortable
#| tbl-cap: Attempts to Evaluate Systemic Resiliency
readxl::read_xlsx("images/author_summary.xlsx") %>%
  kbl(booktabs = TRUE) %>%
  kable_styling()
```

The purpose of transportation networks is to connect locations to each other;
presumably damage to a network would diminish the network's *connectivity*, or
the number of paths between node pairs. It may even leave nodes or groups of
nodes completely isolated. In studies using connectivity as the primary
performance measure, researchers typically apply methods and concepts from graph
theory. These measures may include elementary measures such as the isolation of
nodes in a network [@abdel2007]. More advanced measures have included
heirarchical clustering of node paths [@agarwal2011; @zhang2015], a count of
independent paths [@vodak2019], the reduction of total network capacity
[@ip2011], and special applications of the knapsack and traveling salesman
problems [@guze2014; @osei2014]. Though useful from a theoretical perspective,
many of these authors reported that their approaches tend to break down to some
degree on large, real-world networks where the number of nodes and links numbers
in the tens of thousands, and the degree of connectivity between any arbitrary
node pair is high. They also do not typically account for how network users may
react to the new topology or capacity constraints of the degraded network.

Highway system network failures --- in most imaginable cases --- degrade the
shortest or least cost path, but typically do not eliminate all paths. The
degree to which travel time increases when a particular link is damaged,
however, could provide an estimate of the criticality of that link or node. This
general method has been used to evaluate potential choke points in various
networks [@berdica2007; @jaller2015; @ganin2017] as well as the allocation of
emergency resources [@peeta2010]. Though many applications only consider the
increase in travel time, some authors consider how the users of the network will
respond to the decreased capacity [@ibrahim2011; @serulle2011; @xu2015], and
others attempt to model a shift in departure time or mode [@omer2013].

A primary limitation with increased travel time methodologies is that they
ignore other possible ways a population might adapt its travel to a damaged
network. Aside from shifting routes and modes, people may choose other
destinations and it is possible that some previously planned trips might be
canceled entirely. Travel time-based methods do not account for the costs of
these changes in plans. But accessibility methods --- in particular the
accessibility calculations embedded within many existing regional transport
models ---- provide a framework for evaluating these costs [@ben-akiva1985;
@geurs2004].

Accessibility is an abstract concept with multiple methods of quantification
[@handy1997]. Perhaps the most popular method is the cumulative opportunities
measure: e.g., the number of jobs within a specified travel time threshold. This
is the method employed by Xie and Levinson [@xie2011] in an analysis of the
impact of the I-35 W bridge collapse in Minneapolis. But cumulative
opportunities measures require the analyst to assert a travel time threshold, a
mode, and an opportunity of interest. Some of these assumptions can be relaxed
with a gravity-style access model, but the logsum of a destination choice model
has several benefits as an accessibility term including its grounding in choice
framework, ability to weigh multiple attributes of an alternative, and include
travel impedances by all modes [@dong2006]. These measures can even be weighted
by a cost coefficient to translate the lost utility in monetary terms
[@geurs2004; @geurs2010].

Logsum-derived accessibility meaures have been used before to evaluate network
resiliency [@taylor2008; @masiero2012; @winkler2016]. Where these previous
efforts have been somewhat limited is in their theoretical nature. Though
previous researchers have shown that logsum-derived accessibility measures are
feasible and informative, their use in actual resilience analysis efforts by
state departments of transportation and other relevant agencies appears limited.
