# Results {#results}
In this chapter, we summarize the methodology used to identify
vulnerable links on Utah’s highway network. We also apply the model to scenarios
where critical highway links are removed from the model network. This includes
first a detailed analysis of a single scenario, where I-80 between Salt Lake and
Tooele Counties is severed. We compare the model output to an alternative method
that measures only the change in travel time and does not allow for mode or
destination choice. The model was then applied to 41 individual link closure
scenarios throughout the state.

## Vulnerable Link Identification
Two methodologies were developed to identify vulnerable network links in Utah.
The first method resembles the process used by AEM to determine threat
categories, and threat proximity thresholds to highway links. This methodology
was ultimately not used in the resiliency model. The second method uses an
online Risk Priority Analysis map created by UDOT combined with familiar
knowledge of Utah’s road system. Using this second method, links were identified
due to their location in relation populations or geography, are suspected choke
points, or because they were of interest.


## Single Scenario Analysis
This section outlines an in-depth analysis that was conducted to ensure the
resiliency model was accurately capturing trips with OD pairs in the targeted
area around a closed link. This analysis was done on a link between Tooele and
Salt Lake City, Utah.

Scenario 50, located along I-80 between Tooele and Salt Lake Counties, was
examined to ensure the resiliency model was capturing trips in the vicinity of a
broken link. Analyzing the model outputs at a localized level was necessary to
ensure that the model was appropriately estimating trips and capturing them in
the areas that were expected. In scenario 50, if many effected trips had not
been captured in Tooele and Salt Lake counties, that would have signaled that a
problem was present with the model. This analysis shows that a broken link
mainly effects the area surrounding that link, and the key takeaway here is that
the model functions as intended. Another method to estimate trip costs is the
travel time method, which serves to capture trips that have more fixed OD pairs,
such as freight trips. This analysis also looks at this method in addition to
the logsum estimates.

Table \@ref(tab:tooeletable) compares the overall costs between the logsum and
travel time methods, the specific cost for trips originating in Tooele and
ending in Salt Lake and includes a trip comparison as well. From Table
\@ref(tab:tooeletable), we can see that the logsum method captures about
\$173,996 in experienced expense due to the closure of the link. Specifically,
between Tooele and SLC, the logsum based resiliency model captures \$167,424 of
expense, which is approximately 99% of the total expense as seen in Table
\@ref(tab:tooeletable2). This shows that the resiliency model is effectively
capturing trips in the correct areas based on the percent comparison capture
rate in Table \@ref(tab:tooeletable2). When we look at the travel time method of
analysis, we can see that the costs at both the local and statewide levels are
much greater with \$437,401 and \$406,899 respectively estimated as the costs
due to just the increase in travel time, not using the logit-based model.

```{r tooeletable, echo = FALSE}
read_xlsx("images/tooeletable.xlsx") %>%
  kbl(caption = "Tooele Table", booktabs = TRUE) %>%
  kable_styling() 
```

```{r tooeletable2, echo = FALSE}
read_xlsx("images/tooele_table2.xlsx") %>%
  kbl(caption = "Tooele Table Again", booktabs = TRUE) %>%
  kable_styling( )
```

```{r tooelemap, fig.cap="Total loss from ", dev = "tikz"}
tar_load(zonal_mapdata)

naf0 <- function(x){ifelse(x == 0, NA, x)}

zonal_mapdata <- zonal_mapdata %>%
  st_transform(4326) %>% 
  mutate(across(c(HBW, HBO, NHB), naf0)) %>%
  filter(!is.na(HBW)) 

if (!knitr::is_latex_output()){
  tazpal <- colorNumeric("Reds", c(zonal_mapdata$HBO, zonal_mapdata$HBW, zonal_mapdata$NHB))
  
  leaflet( zonal_mapdata ) %>%
    addProviderTiles(providers$Esri.WorldGrayCanvas) %>%
    addPolygons(color = ~tazpal(HBW), stroke = 0)
} else {
  pbox <- st_bbox(zonal_mapdata)
  
  # overall map
  map_back <- get_map(#c(lon = mean(pbox[c(1, 3)]), lat = mean(pbox[c(2, 4)]) ),  
                      c(-112.22547310384623, 40.71921168912544),
                      zoom = 10, source = "google", maptype = "roadmap", color = "bw")
  
  ggmap(map_back, extent = "device") + 
    coord_sf(crs = st_crs(4326),  expand = FALSE) + 
    geom_sf(data = zonal_mapdata , inherit.aes = FALSE, aes(fill = HBW), 
            alpha = 0.5, lwd = 0)  + 
    theme(axis.line = element_line(color = NA))  + 
    scale_fill_viridis("Total HBW Utility Loss", option = "magma", direction = 1) + 
    xlab("") + ylab("") 
  
}

```


The logsum and travel time methods for determining the overall cost associated
with can be broken down into overall costs, and comparable costs. The comparable
costs are made up of those purposes which are included in the resiliency model.
HBW, HBO, and NHB trips are included in both models, thus the results from both
the logsum model and the travel time method can be compared to each other.

The travel time method measures the difference in travel time between the base
scenario and any other scenario, and then multiplies that difference by the VOT
for each trip purpose and the number of trips estimated for each trip purpose.
For external trips, freight trips, and REC trips, these were all extracted
directly from USTM. The HBW, HBO, and NHB purposes are estimated using the
logsum model. Calculating the costs associated with the difference in travel
time allows a better estimation of the true cost experienced by road users who
are not included in the logsum based resiliency model. So, while the logit-based
model is more sensitive to changes in the network, it is not able to truly
estimate the costs associated with closure for those purposes that are not
included in the analysis.

## Comparative Scenario Results
We now apply the model to compare 40 additional scenarios where individual
highway facilities are removed from the model highway network. These scenario
locations are shown in Figure \@ref(fig:links), and were identified in a report
by AEM (2017). AEM approached the idea of systemic resiliency by attempting to
classify various types of threats toward specific infrastructure types. This
method, while valid, was difficult to implement statewide. Some facilities have
obvious natural threats such as earthquake, flood, or landslide.

```{r links-map, fig.cap="Total cost of link closure.", fig.show='hold', dev = "tikz"}
tar_load(mapdata)


if(!knitr::is_latex_output()){
  
  pal <- colorNumeric("Reds", mapdata$value[!is.na(mapdata$value)]) 
  
  leaflet(mapdata %>% st_transform(4326)) %>%
    addProviderTiles(providers$Esri.WorldGrayCanvas) %>%
    addCircleMarkers(color = ~pal(value))
    
} else {
  pbox <- st_bbox(mapdata)
  
  # overall map
  map_back <- get_map(c(lon = mean(pbox[c(1, 3)]), lat = mean(pbox[c(2, 4)]) ),  
                      zoom = 7, source = "google", maptype = "roadmap", color = "bw")
  
  ggmap(map_back, extent = "device") + 
    coord_sf(crs = st_crs(4326),  expand = FALSE) + 
    geom_sf(data = mapdata, inherit.aes = FALSE, size = 5, stroke = 1, shape = 21,
            aes(fill = value), alpha = 0.5)  + 
    theme(axis.line = element_line(color = NA))  + 
    scale_fill_viridis("Cost of closure (\\$)", option = "magma", direction = -1) + 
    xlab("") + ylab("") 
  
  
  # SLC map
  
  
}
```


This section will present the results of the 41 scenarios analyzed. First, Table
\@ref(tab:linktable) shows each of the scenarios, labeled “road10” for scenario
10, and “road11” for scenario 11 along with route numbers or street names, as
well as a physical location where the link was cut.

```{r linktable, echo = FALSE}
read_xlsx("images/linktable.xlsx") %>%
  kbl(caption = "Link Table", booktabs = TRUE) %>%
  kable_styling()
```

The logsum method results are as follows in Table 14. The results are ranked
from the road with the largest cost to the road with the smallest cost. We can
see that road 50, which corresponds to I-80 between SLC and Tooele, experiences
the largest cost per day according to the resiliency model. Following road 50,
roads 38, 27, 37, and 14 make up the five most important road according to the
cost estimation provided by the resiliency model. Each of these roads is an
Interstate facility in Northern Utah, which is heavily populated. Some roads,
such as road 10, which is along SR-95 near Hite, experiences no measurable
change to HBW, HBO, or NHB traffic. This is likely due to the remote location of
the highway link that was cut. A ranking is provided for all of the roads in
Table \@ref(fig:logsumrank).

```{r logsumrank, echo = FALSE}
read_xlsx("images/logsumrank.xlsx") %>%
  kbl(booktabs = TRUE, caption = "Total cost in dollars") %>%
  kable_styling()
```

Table \@ref(fig:traveltimerank) contains the ranking results from the travel
time method. Here, we see that the first five roads differ from the results of
the logsum model. Instead, road 48 becomes the most important road due to costs
experienced. Road 48 is a part of I-70 near Green River, Utah. The other four
roads that make up the top five most important roads in the travel time method
analysis are road 13, road 20, road 50, and road 49. Some of these roads appear
in both the logsum and travel time methods. Here again, several of the roads
that are most important are located in Northern Utah. It is important to note
that the main driving factor as to why a road was important or not in the travel
time analysis was how much freight and external traffic it experienced.
Including the freight, even with the logsum results, changes the rankings
drastically.

```{r traveltimerank, echo = FALSE}
read_xlsx("images/traveltimerank.xlsx") %>% 
  kbl(booktabs = TRUE, caption = "Total cost in dollars") %>%
  kable_styling()
```

Some other interesting findings are that in the top 10 of each analysis methods,
three scenarios appear in both rankings. Road 50, which corresponds to I-80
between Tooele and SLC, road 27 which corresponds to I-84 in Weber Canyon, and
road 37 which corresponds to I-80 in Parley’s Canyon are included in the top 10
scenarios for both methods of analysis. This is likely due to the number of
passenger trips along these routes and the number of freight trips that occur
along these routes.

## Positive Benefit Scenarios
Five of the scenarios indicated a benefit resulting from highway link closure.
These scenarios were examined more closely to determine what causes these
atypical results. The affected links are all located in Salt Lake Valley at the
following locations: Bangerter Highway near Bluffdale, I-80 near 1300 E, I-15
between 2100 S and 1300 S, I-215 near Cottonwood Heights, and Mountain View
Corridor near West Jordan.

It was determined that a likely cause which explains these atypical results is
that the shortest path by time is not the shortest path by distance. The
automobile accessibility is determined by the AM congested travel time in the
Utah Statewide Travel Model (USTM). The travel distance – used to determine the
accessibility of destinations by walking – is the distance of that path, and not
the actual shortest distance path as might be more preferable.

When a highway link is broken, the new shortest path by time is longer than in
the base scenario with this link available. But the new path may actually be
shorter by distance. This causes an increase in the utility of accessing
destinations by non-motorized modes, potentially overwhelming the decrease in
automobile utility.

This is only observed in heavily urbanized regions for two reasons:  

-	The presence of high-speed expressways and parallel local roads means that
alternate paths with shorter distances but longer vehicle times are more likely.

-	The increased availability of destinations within the non-motorized distance
threshold (50 miles) means that alternative destinations exist.

Overall, the results of the analysis indicate that the likely cause of a
positive cost being estimated for these five scenarios is that there are easily
accessible alternate routes in the area, along with competing TAZ of similar
size in the DC size term equation.

## Summary
The overall results show that the resiliency model is more sensitive to network
changes than the travel time comparison. The ability for a user to choose both a
mode and destination (or alternate destination) cause the logsum results to
often estimate a smaller cost than the travel time results would. However, when
the travel time results are factored in, the overall rankings of the 41
scenarios considered change dramatically. This is due to the large expenses
experienced by freight traffic, which has a much higher VOT than other passenger
trips do. In summary, Table \@ref(fig:logsumrank) and Table
\@ref(fig:traveltimerank) show the rankings for both the logsum and travel time
analysis respectively. The logsum suggests that I-80 between Tooele and SLC is
the most important road, while the travel time method, or total priority,
indicates that I-70 near Green River is the most important road due to cost
associated with closure.





